---
title: "Data Cleaning"
author: "Elyse McFalls"
date: "2024-02-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F)
```

```{r}
library(tidyverse)
library("readxl")
```

Current data issues
- no file for demographic info by school
- the CTE files have a different number of schools reprenseted (the file on CTE concentrators (2018-2022 data) only has 791 schools while the file on cte enrollment clusters (only 2019 data) has 1,113 schools and the file on cte enrollment (2010-2022 data) in general has 1,615 schools)
- the CTE cluster enrollment file is really unclear: are those the clusters that were offered at each school?


# Loading Data

```{r}
concentrators <- read.csv("data/2017-2022 CTE_concentrators.xlsx - Sheet1.csv")
absenteeism <- read.csv("data/chronic_absent.csv")
tsi_csi <- read_excel("data/tsi_and_csi.xlsx")
#standards <- read.csv("data/rcd_acc_pc (1).txt", sep = "\t")
```

```{r}
cte_enrollment <- read_excel("data/rcd_cte_enrollment.xlsx")
demographics <- read_excel("data/rcd_prin_demo.xlsx")
school_info <- read_excel("data/rcd_location.xlsx")
cte_enrollment_clusters <- read_excel("data/rcd_cte_enrollment_cluster.xlsx")
```

# Cleaning Data


Only keeping data from 2018-2022

```{r}
years <- c(2018, 2019, 2020, 2021, 2022)
```

```{r}
concentrators <- concentrators %>%
  filter(year %in% years)

tsi_csi <- tsi_csi %>%
  filter(year %in% years)

school_info <- school_info %>%
  mutate(title_i = ifelse(is.na(title_i), 'N', title_i)) %>%
  filter(year %in% years)
```


```{r}
demo <- demographics %>%
  filter(year %in% years) %>%
  complete(year, agency_code, subgroup, fill = list(pct_prin_demo = 0)) %>%
  pivot_wider(names_from = subgroup, values_from = pct_prin_demo) %>%
  mutate(majority_white = ifelse(WH7 > 0.5, T, F))
```

```{r}
# widening concentrators data
concentrators <- concentrators %>%
  complete(year, agency_code, career_cluster, fill = list(num_concentrators = 0)) %>%
  pivot_wider(names_from = career_cluster, values_from = num_concentrators) 
```

```{r}
# getting data for clusters per school
clusters_per_school <- cte_enrollment_clusters %>%
  mutate(i = 1) %>%
  complete(year, agency_code, career_cluster, fill = list(i = 0)) %>%
  pivot_wider(names_from = career_cluster, values_from = i)
```

```{r}
absenteeism <- absenteeism %>%
  mutate(agency_code = paste0(0, agency_code)) %>%
  select(-count, -den, -masking) %>%
  pivot_wider(names_from = subgroup, values_from = pct, names_prefix = "absent_group_")
```

# Combining data

```{r}
school_info <- school_info %>%
  # only select schools with CTE enrollment data
  filter(agency_code %in% cte_enrollment$agency_code) %>%
  left_join(cte_enrollment, by = c("year", "agency_code")) %>%
  left_join(tsi_csi, by = c("year", "agency_code")) %>%
  left_join(demo, by = c("year", "lea_code" = "agency_code")) %>%
  left_join(absenteeism, by = c("year", "agency_code"))
```

```{r}
 df_concentrator <- school_info %>%
  # only select schools with CTE enrollment data
  filter(agency_code %in% cte_enrollment$agency_code) %>%
  #mutate(title_i = ifelse(is.na(title_i), 'N', 'Y')) %>%
  #left_join(cte_enrollment, by = c("year", "agency_code")) %>%
  left_join(concentrators, by = c("year", "agency_code")) %>%
  select(-street_addr, -street_addr2, -state, -phone, -url, -super_name, -esea, -stem) %>%
  mutate_at(vars(AAVC:TRAN), funs(round((. / stu_enroll) * 100, 3)))
```

```{r}
clusters_per_school <- school_info %>%
  filter(year == 2019) %>%
  # only select schools with CTE enrollment data
  filter(agency_code %in% cte_enrollment$agency_code) %>%
  #mutate(title_i = ifelse(is.na(title_i), 'N', 'Y')) %>%
  left_join(clusters_per_school, by = c("year", "agency_code"))
```

```{r}
write.csv(school_info, "data/school_information.csv")
write.csv(df_concentrator, "data/concentrator_information.csv")
write.csv(clusters_per_school, "data/career_cluster_information.csv")
```


# EDA

## Absenteeism

```{r}
school_info %>%
  filter(year %in% c(2018:2020)) %>%
  group_by(title_i) %>%
  summarise(mean_pct_absent = mean(absent_group_ALL, na.rm = T))
```

```{r}
school_info %>%
  filter(year %in% c(2018:2020)) %>%
  group_by(majority_white) %>%
  summarise(mean_pct_absent = mean(absent_group_ALL, na.rm = T))
```

```{r}
# consistently underperforming subgroups
school_info %>%
  filter(year %in% c(2018:2020)) %>%
  group_by(tsi_cu) %>%
  summarise(mean_pct_absent = mean(absent_group_ALL, na.rm = T))
```


```{r}
# consistently underperforming subgroups
school_info %>%
  filter(year %in% c(2018:2020)) %>%
  group_by(title_i, tsi_cu) %>%
  count()
```

```{r}
# low performing 
school_info %>%
  filter(year %in% c(2018:2020)) %>%
  group_by(csi_lp) %>%
  summarise(mean_pct_absent = mean(absent_group_ALL, na.rm = T))
```

## Are there differences in what clusters are available?

```{r}
# all 2019 data, except for CD, non title I schools have a higher percentage of each cluster 
# plot absolute difference 
clusters_per_school %>%
  group_by(title_i) %>%
  summarise(across(AAVC:TRAN, ~ mean(.x, na.rm = TRUE))) %>%
  pivot_longer(!title_i, names_to = "career_cluster", values_to = "pct_have") %>%
  ggplot(aes(x = career_cluster, y = pct_have, fill = title_i, group = title_i)) +
  geom_bar(stat = "identity", position = "dodge")
```

```{r}
# not title 1 - title 1
clusters_per_school %>%
  group_by(title_i) %>%
  summarise(across(AAVC:TRAN, ~ mean(.x, na.rm = TRUE))) %>%
  pivot_longer(!title_i, names_to = "career_cluster", values_to = "pct_have") %>%
  arrange(career_cluster) %>%
  ungroup () %>%
  group_by(career_cluster) %>%
  summarise(access_disparity = pct_have[1] - pct_have[2]) %>%
  ggplot(aes(x = career_cluster, y = access_disparity)) +
  geom_bar(stat = "identity") +
  geom_text(position = "top", )
```

```{r}
# majrotity white schools - majority non-white schools 
clusters_per_school %>%
  group_by(year, majority_white) %>%
  summarise(across(AAVC:TRAN, ~ mean(.x, na.rm = TRUE))) %>%
  pivot_longer(!c(year, majority_white), names_to = "career_cluster", values_to = "pct_have") %>%
  arrange(career_cluster) 
```


```{r}
clusters_per_school %>%
  group_by(majority_white) %>%
  summarise(across(AAVC:TRAN, ~ mean(.x, na.rm = TRUE))) %>%
  pivot_longer(!majority_white, names_to = "career_cluster", values_to = "pct_have") %>%
  ggplot(aes(x = career_cluster, y = pct_have, fill = majority_white, group = majority_white)) +
  geom_bar(stat = "identity", position = "dodge")
```

```{r}
clusters_per_school %>%
  group_by(majority_white) %>%
  summarise(across(AAVC:TRAN, ~ mean(.x, na.rm = TRUE))) %>%
  pivot_longer(!majority_white, names_to = "career_cluster", values_to = "pct_have") %>%
  ggplot(aes(x = career_cluster, y = pct_have, fill = majority_white, group = majority_white)) +
  geom_bar(stat = "identity", position = "dodge")
```


```{r}
# each school type has the same top 5 common clusters, although they are more prevelant in non title i schools 
clusters_per_school %>%
  group_by(title_i) %>%
  summarise(across(AAVC:TRAN, ~ mean(.x, na.rm = TRUE)))
```
```{r}
# each school type has the same top 5 common clusters, although they are more prevelant in non title i schools 
clusters_per_school %>%
  group_by(majority_white) %>%
  summarise(across(AAVC:TRAN, ~ mean(.x, na.rm = TRUE)))
```

## Are there differences in CTE participation?

```{r}
# participation rates were similar from 2018-2021 then dropped off in 2022
df_concentrator %>%
  group_by(year, title_i) %>%
  summarise(mean_pct = mean(pct, na.rm = T)) %>%
  ggplot(aes(x =year, y = mean_pct, group = title_i, color = title_i)) +
  geom_line()
```

```{r}
# participation rates were similar from 2018-2021 then dropped off in 2022
df_concentrator %>%
  group_by(year, majority_white) %>%
  summarise(mean_pct = mean(pct, na.rm = T)) %>%
  pivot_wider(names_from = year, values_from = mean_pct)
```


## What schools do not have cluster concentrator data?

```{r}
# title 1 schools are less likely to have data for concentrators
df_concentrators %>%
  mutate(no_concen_data = ifelse(is.na(EDUC), T, F)) %>%
  group_by(title_i, no_concen_data) %>%
  count() %>%
  group_by(title_i) %>%
  mutate(prop = n/sum(n))
```

```{r}
# interestingly half of the CTE schools do not have CTE data
# vocational, alternative, and magnet schools are more likely to have this data
df_concentrators %>%
  mutate(no_concen_data = ifelse(is.na(EDUC), T, F)) %>%
  group_by(school_type, no_concen_data) %>%
  count() %>%
  group_by(school_type) %>%
  mutate(prop = n/sum(n))
```




## Where do most kids concentrate for each school type

```{r fig.width=10, fig.height=4.5}
# it looks like the clusters most students concentrate in is pretty simialr for both school types
df_concentrator %>%
  group_by(year, title_i) %>%
  summarise(across(AAVC:TRAN, ~ mean(.x, na.rm = TRUE))) %>%
  pivot_longer(!c(year, title_i), names_to = "career_clusters", values_to = "avg_pct_concentrator") %>%
  ggplot(aes(x = year, y = avg_pct_concentrator, group = career_clusters, color = career_clusters)) +
  geom_line() +
  geom_point() +
  facet_wrap(~title_i)
```

```{r}
df %>%
  group_by(year, title_i) %>%
  summarise(across(AAVC:TRAN, ~ mean(.x, na.rm = TRUE))) %>%
  filter(title_i == 'N') %>%
  summarise(across(AAVC:TRAN, ~ max(.x, na.rm = TRUE)))
```

```{r}
df %>%
  group_by(year, title_i) %>%
  summarise(across(AAVC:TRAN, ~ mean(.x, na.rm = TRUE))) %>%
  filter(title_i == 'Y') %>%
  summarise(across(AAVC:TRAN, ~ max(.x, na.rm = TRUE)))
```